### 2022-03-18 tex/syntax.tex
[ backport ] syntax from ESOP22 to thesis

We take the insights of the previous section and use them to build a
generic framework for posemiring-annotated substructural systems in
Agda. We will first show \emph{descriptions} of systems, which are
comprised of rules that have premises combined using the bunched
combinators. We then show how to construct the Agda data type of
intrinsically well scoped, typed, and resourced terms for any given
system of our framework. We use the prototypical system from
\cref{fig:lr-comb} as a running example. \cref{sec:other-syntaxes}
presents further examples that our framework can express.

@@

\paragraph{\cref{fig:lr-comb} as a \AgdaRecord{System}.}

@@

\paragraph{Terms of a \AgdaRecord{System}.}

@@

With the first clause of \AgdaFunction{$\llbracket$\_$\rrbracket$p} explained,
the rest are simply interpretations of premises into bunched combinators.

@@

We use \AgdaFunction{Scope} to deal with new variables in syntax.
Terms resemble the free monad over a layer-of-syntax functor, though
that picture is complicated by variable binding.  A term is either a
variable or a use of a logical rule together with terms for each of
the required subterms. The \AgdaFunction{Size} argument is where we
use sized types to convince Agda that this type is strictly positive.

@@

Terms defined like this are still quite difficult to write, mainly because of
frequently changing usage contexts and the need for proofs that they all match
up.
We will see how to automate these proofs in \cref{sec:usage-elaborator}.

>>>>

We take the insights of the previous section and use them to build a
generic framework for posemiring-annotated substructural systems in
Agda. We will first show \emph{descriptions} of systems, which are
comprised of rules that have premises combined using the bunched
combinators. We then show how to construct the Agda data type of
intrinsically well scoped, typed, and resourced terms for any given
system of our framework. We use the prototypical system from
\cref{fig:lr-comb} as a running example. \Cref{sec:other-syntaxes}
presents further examples that our framework can express.

@@

% \paragraph{\Cref{fig:lr-comb} as a \AgdaRecord{System}.}

@@

\subsection{Terms of a \AgdaRecord{System}}\label{sec:terms}

@@

The remainder of the clauses of \AgdaFunction{$\llbracket$\_$\rrbracket$p}
are the interpretation into bunched combinators.
The superscript \AgdaFunction{$^c$} on the bunched connectives denotes that
they have been lifted from predicates on usage vectors to predicates on
contexts, with the type component of the context shared throughout.
Additive connectives \AgdaFunction{$\dot1$} and \AgdaFunction{$\dottimes$} are
already polymorphic (not relying on anything specific about usage vectors), so
do not need a \AgdaFunction{$^c$} variant.

@@

We use \AgdaFunction{Scope} to deal with new variables in syntax.
Terms resemble the free monad over a layer-of-syntax functor, though
that picture is complicated by variable binding.  A term is either a
variable or a use of a logical rule together with terms for each of
the required subterms. The \AgdaFunction{Size} argument is a use of
Agda's sized types to record that subterms are smaller than the
surrounding term for the termination checker.

@@

This definition uses \AgdaFunction{$\dotto$}, which, analogously to
\AgdaFunction{$\dottimes$}, is an index-preserving version of the function
space.
We take \AgdaFunction{$\dotto$} to handle $n$ many indices --- in this
case two (the context and the type).
The notation
\AgdaFunction{$\forall[$}\AgdaSpace{}\AgdaBound{T}\AgdaSpace{}\AgdaFunction{]}
stands for
\AgdaSymbol{$\forall$}\AgdaSpace{}\AgdaSymbol{\{}%
\AgdaBound{x$_1$}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaBound{x$_n$}%
\AgdaSymbol{\}}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaBound{T}%
\AgdaSpace{}%
\AgdaBound{x$_1$}\AgdaSpace{}\AgdaSymbol{$\ldots$}\AgdaSpace{}\AgdaBound{x$_n$},
where \AgdaBound{T} is a type family with $n$ many indices.

Terms in this data type are difficult to write by hand, due to the
need for proofs that the usage contexts are handled correctly. For
example, the following term is needed to show that, in the $\{\gr0,
\gr1, \gr\omega\}$ (linearity) posemiring of \cref{thm:linearity},
$\oc\gr\omega$ forms a comonad.
Pattern synonyms \AgdaInductiveConstructor{$\multimap$I},
\AgdaInductiveConstructor{!E$'$}, and
\AgdaInductiveConstructor{!I$'$} stand for applications of
\AgdaInductiveConstructor{`con}, with the latter two taking explicit usage
contexts and proofs.
On concrete
posemirings (as in this example), unification is particularly poor at
inferring the usage contexts from the proofs because addition and
multiplication are no longer (judgementally) injective.
The function \AgdaFunction{var\#} is a way of turning a statically known de
Bruijn level and a usage proof into an application of \AgdaInductiveConstructor{`var}.

\ExecuteMetaData[\HeavyItex]{cojoin-explicit}

 Writing terms like this
is clearly unsustainable. We will see a way of automating the
necessary proofs via a \AgdaRecord{System}-generic elaborator in
\cref{sec:usage-elaborator}.
